A non-contextual hidden variable model for quantum mechanics

### 2016-07-11 finalPaper.tex
made various changes

%% LyX 2.1.4 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[british,aps,prl,superscriptaddress,nofootinbib,times,reprint]{revtex4-1}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm,headheight=2cm,headsep=2cm,footskip=2cm}
\setcounter{secnumdepth}{3}
\usepackage{verbatim}
\usepackage{refstyle}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{esint}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.

\AtBeginDocument{\providecommand\thmref[1]{\ref{thm:#1}}}
\AtBeginDocument{\providecommand\lemref[1]{\ref{lem:#1}}}
\AtBeginDocument{\providecommand\propref[1]{\ref{prop:#1}}}
\AtBeginDocument{\providecommand\eqref[1]{\ref{eq:#1}}}
\AtBeginDocument{\providecommand\figref[1]{\ref{fig:#1}}}
\AtBeginDocument{\providecommand\secref[1]{\ref{sec:#1}}}
\AtBeginDocument{\providecommand\ssecref[1]{\ref{ssec:#1}}}

\newref{sec}{name=Section~}
\newref{ssec}{name=Subsection~}

@@

\usepackage{babel}
  \providecommand{\definitionname}{Definition}
  \providecommand{\examplename}{Example}
  \providecommand{\lemmaname}{Lemma}
  \providecommand{\propositionname}{Proposition}
  \providecommand{\remarkname}{Remark}
  \providecommand{\theoremname}{Theorem}
  \providecommand{\notationname}{Notation}

\begin{document}

\title{A non-contextual hidden variable model for Quantum Mechanics}

\author{Atul Singh Arora}

\email{ms11003@iisermohali.ac.in}

\selectlanguage{british}%

\affiliation{Indian Institute of Science Education and Research (IISER), Mohali}

\author{Arvind}

\email{arvind@iisermohali.ac.in}

\selectlanguage{british}%

\affiliation{Indian Institute of Science Education and Research (IISER), Mohali}
\begin{abstract}
We propose a non-contextual hidden variable model, consistent with all predictions of Quantum Mechanics (QM). We reconsider the no-go theorems on hidden variables and conclude that the notion of contextuality is not a necessary feature of QM. Motivated by consistency requirements between Bohmian Mechanics and the said theorems, we provide an alternative view. This is accomplished by identifying a new classical property, we call `multiplicativity'. It turns out that if this condition is relaxed one is able to construct hidden variable models which are consistent with QM. Advantages of this view are illustrated by considering the implications of non-multiplicativity to non-locality. Our model is for finite dimensional quantum systems, however, generalization to the infinite dimensional case should be possible. %and its relation to contextuality.
\end{abstract}
\maketitle

\section{Introduction}

Einstein's \cite{EinsteinEPR} work on incompleteness of QM and the subsequent seminal work of Bell \cite{Bell1964}, assessing compatibility of hidden variable (HV) models with QM has proven to be repeatedly useful. In Bell's case, for instance, it gave a precise meaning to the foundational notion of locality. Further, it has been of pragmatic utility in the areas of security, randomness certification etc. \cite{Ekert,PironioRndmnssCrtfcn} While Bell showed that local HV models can not describe nature, the Kochen Specker theorem \cite{KochenSpecker} (and related work \cite{Gleason,BellOnHiddenVariables,Peres,Mermin}) ruled out an even larger class of HV models. Consequently the notion of contextuality was identified, methods to experimentally test it were proposed and evidence in its favour reported \cite{SimonContExpProp,HuangContExp,YangContExp,HasegawaContExp}. Contextuality also, in recent times has been harnessed for computation and cryptography \cite{HowardCntxCmptn,CabelloCntxScrt}. []Contextuality has also stirred further investigation of foundational aspects of QM: relation between contextuality and classicality has been explored \cite{PawelCntxClsscl}, contextuality has been quantified in terms of memory \cite{MatthiasCntxMmry} and bounded memory in conjunction with contextuality has been used to show completeness of QM \cite{CabelloMmryQM}.[] Not all HV models were, however, proved to be incompatible by the aforesaid theorem. Bohm's work \cite{Bohm1,Bohm2} is one such (in)famous example. This approach has allowed for an interpretation of QM with precise trajectories.

%Previous work {[}the essentials I have already discussed{]}: 
We take a closer look at the aforesaid no-go theorems and find that contextuality is in fact not a necessary feature of QM (see \figref{block}). This has been shown earlier by taking disturbances due to measurements into consideration \cite{NoContextuality,LaCourNoCntx}. In this paper, we construct a significantly simpler toy model for illustrating not only the aforesaid but to also delineate a complete alternative to contextuality. The toy model is a specific case of what we call a c-ingle variable theory. Its predictions match with those of QM for systems described by arbitrary finite dimensional Hilbert spaces.

%Triangle: <TODO: ask arvind sir, see \figref{block}>

\begin{figure}[h]
\includegraphics[width=\columnwidth]{block1}
\caption{Non-contextuality is not inconsistent with QM}
\label{fig:block}\end{figure}

Organization of this paper is as follows. In \secref{context} we establish a notation and definitions to facilitate further discussions. In \secref{nonMultiplicativity} we introduce the notion of multiplicativity and that of sequential multiplicativity. We review the `contextuality' theorem and prove some results about multiplicativity. We construct an explicit non-contextual model in \ssecref{cingle} and apply it to the very proof of `contextuality' to illustrate non-multiplicativity. This is supplemented by a discussion about Bohmian Mechanics in \ssecref{nonContBohm} where contextuality is removed by hand. The paper is concluded with discussion of implications of multiplicativity on locality, entanglement, superposition and apparent contextuality in \secref{implications}.

\section{Context \label{sec:context}}

The notation used henceforth has been defined below.

\begin{notn} (a) $\psi\in\mathcal{H}$ would typically represent the quantum mechanical state of the system (assumed pure), residing in the Hilbert space $\mathcal{H}$, (b) $\hat{\mathcal{H}}$ is defined to mean $\mathcal{H}\otimes\mathcal{H}^{\dagger}$ (i.e. the set of Hermitian operators, observables, on the space $\mathcal{H}$), (c) $[\mathcal{H}]$ is defined to mean $(\mathcal{H},\,\mathbb{R}^{\otimes})$, which represents the state of the system including HVs, (d) $[\psi]\in[\mathcal{H}]$ will represent the state of the system, including HVs, (e) a prediction map is $m:\hat{\mathcal{H}},[\mathcal{H}]\to\mathbb{R}$, (f) a sequence map is $s:\hat{\mathcal{H}},[\mathcal{H}],\mathbb{R}\to[\mathcal{H}]$, (g) $f$ is an arbitrary map from $\{ \hat{\mathcal{H}},\hat{\mathcal{H}},\dots\hat{\mathcal{H}} \} \to \hat{\mathcal{H}}$ constructed using multiplication and addition of the observables, and multiplication with complex numbers\footnote{Strictly, the map $f$ would depend on the operators to ensure preservation of Hermiticity; thus a given $f$ will be defined only for a subset of $\{ \hat{\mathcal{H}},\hat{\mathcal{H}},\dots\hat{\mathcal{H}} \}$.}, (h) $\tilde{f}$ is a map constructed by replacing observables in $f$ with real numbers. 
%\footnote{$\tilde f$ constructed this way should be written as $\tilde f:\mathbb{R}^{\otimes N} \to \mathbb{C}$. In fact the values should be restricted to the spectrum of the associated observables.}.
\end{notn}

In addition to the notation, we use the following definition of non-contextuality.
\begin{defn}
A theory is non-contextual, if it provides a map $m: \{ \hat{\mathcal{H}},[\mathcal{H}] \} \to\mathbb{R}$ to explain measurement outcomes. A theory which is not non-contextual is contextual.\end{defn}
\begin{rem}
The notion of non-contextuality may be extended to maps. A map is non-contextual if it is of the form $m: \{ \hat{\mathcal{H}},[\mathcal{H}] \} \to\mathbb{R}$; prediction maps are non-contextual.
\end{rem}

In the literature broader definitions have been suggested which declare a larger set of theories as non-contextual. For our purposes however, this restricted definition will suffice. The term contextual is used to suggest that the value an operator takes might depend on which other compatible observable it is measured with.

\section{Non-Multiplicativity: An alternative to Contextuality \label{sec:nonMultiplicativity}}

One can show \cite{KochenSpecker} that under reasonable assumptions, no non-contextual theory can explain all predictions of QM. We re-state these results and show that nowhere does contextuality appear to be a logical necessity. Instead we note that non-contextual theories with certain properties, we call \emph{multiplicativity} are incompatible with QM. Viewed from this perspective, non-\emph{multiplicativity} is the non-classical alternative to contextuality in QM. Precise definitions of this and a related property \emph{sequential multiplicativity}, are given below.

\begin{defn}
A prediction map $m$ is \emph{multiplicative} iff 
% \[
% \begin{split}
% m(f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{N}),[\psi]) = \\
% f(m(\hat{B}_{1},[\psi]),m(\hat{B}_{2},[\psi]),\dots m(\hat{B}_{N},[\psi])),
% \end{split}
% \]

\begin{align*}
& m(f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{N}),[\psi]) = \\
& \tilde f(m(\hat{B}_{1},[\psi]),m(\hat{B}_{2},[\psi]),\dots m(\hat{B}_{N},[\psi])),
\end{align*}
where $\hat{B}_{i}\in\hat{\mathcal{H}}$ are arbitrary mutually commuting observables and $[\psi]\in[\mathcal{H}]$. A non-\emph{multiplicative} map is one that is not multiplicative.
\end{defn}
Note that if $m$ is taken to represent the measurement outcome (in
QM), then for states of the system which are simultaneous eigenkets
of $\hat{B}_{i}$s, $m$ must clearly be multiplicative. It is, however,
not obvious that this property must always hold. To motivate this,
consider the state of two spin half particles, $\left|11\right\rangle :=\left|1\right\rangle \otimes\left|1\right\rangle $
written in the computational basis, $\hat{B}_{1}=\hat{\sigma}_{x}\otimes\hat{\sigma}_{x}$,
$\hat{B}_{2}=\hat{\sigma}_{y}\otimes\hat{\sigma}_{y}$ and $\hat{C}=\hat{B}_{1}\hat{B}_{2}=-\hat{\sigma}_{z}\otimes\hat{\sigma}_{z}$.
We must have $m(\hat{C})=-1$ while $m(\hat{B}_{1})=\pm1$ and $m(\hat{B}_{2})=\pm1$,
independently, according to QM, with probability half. Here multiplicativity doesn't seem to hold. Antithetically it is clear that if one first measures
$\hat{B}_{1}$ and subsequently measures $\hat{B}_{2}$, then the
product of the results must be $-1$. This is consistent with measuring
$\hat{C}$. To capture this property we define \emph{sequential multiplicativity}
as follows.
\begin{defn}
A prediction map $m$ is \emph{sequentially multiplicative} for a
given sequence map $s$, iff 
\begin{align*}
&m(f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{N}),[\psi_{1}])=\\
&\tilde f(m(\hat{B}_{1},[\psi_{k_{1}}]),m(\hat{B}_{2},[\psi_{k_{2}}]),\dots,m(\hat{B}_{N},[\psi_{k_{N}}])),
\end{align*}
where $\mathbf{k}=(k_{1},k_{2},\dots,k_{N})\in\{(1,2,3\dots N),(2,1,3\dots N)$ + permutations, making $N!$ terms$\}$,
$\hat{B}_{i}\in\hat{\mathcal{H}}$ are arbitrary mutually commuting
observables, $[\psi_{i}]\in[\mathcal{H}]$ and $[\psi_{k+1}]\equiv s(\hat{B}_{k},[\psi_{k}],m(\hat{B}_{k},[\psi_{k}]))$,
$\forall\,[\psi_{i}]$.
\end{defn}
We now discuss the `proofs of contextuality' in the purview of the
aforesaid.
\begin{thm}
Let a map $m:\hat{\mathcal{H}}\to\mathbb{R}$, be s.t. (a) $m(\hat{\mathbb{I}})=1$,
(b) $m(f(\hat{A}_{1},\hat{A}_{2},\dots))=\tilde f(m(\hat{A}_{1}),m(\hat{A}_{2}),\dots)$,
where $\hat{A}_{i}$ are arbitrary Hermitian operators. If $m$ is assumed to describe the outcomes of measurements, then no $m$ exists which is consistent with all predictions of QM\label{thm:det} (see \ref{proofs} for proofs). \end{thm}
Here $m$ maybe viewed as a specific class of prediction maps, that
implicitly depends on the state $[\psi]$. It follows then that the
maps ruled out by the theorem are non-contextual. The set of maps
ruled out, however, can be enlarged by imposing fewer restrictions
on it. This is achieved by the following theorem.

\begin{thm}
Let a map $m:\hat{\mathcal{H}}\to\mathbb{R}$, be s.t. (a) $m(\hat{\mathbb{I}})=1$,
(b) $m(f(\hat{B}_{1},\hat{B}_{2},\dots))=\tilde f(m(\hat{B}_{1}),m(\hat{B}_{2}),\dots)$,
for any arbitrary function $f$, where $\hat{B}_{i}$ are mutually
commuting Hermitian operators. If $m$ is assumed to describe the
outcomes of measurements, then no $m$ exists which is consistent
with all predictions of QM (see \ref{proofs} for more proofs).\label{thm:KS}\end{thm}

% The following proof is not only more general, but also independent
% of the initial state of the system.
\begin{proof}
Peres Mermin ($\left|\mathcal{H}\right|\ge4$) \cite{Peres,Mermin}:

Consider the following set of operators 
\[
\hat{A}_{ij}\doteq\left[\begin{array}{ccc}
\hat{\mathbb{I}}\otimes\hat{\sigma}_{x} & \hat{\sigma}_{x}\otimes\hat{\mathbb{I}} & \hat{\sigma}_{x}\otimes\hat{\sigma}_{x}\\
\hat{\sigma}_{y}\otimes\hat{\mathbb{I}} & \hat{\mathbb{I}}\otimes\hat{\sigma}_{y} & \hat{\sigma}_{y}\otimes\hat{\sigma}_{y}\\
\hat{\sigma}_{y}\otimes\hat{\sigma}_{x} & \hat{\sigma}_{x}\otimes\hat{\sigma}_{y} & \hat{\sigma}_{z}\otimes\hat{\sigma}_{z}
\end{array}\right]
\]
which have the property that all operators along a row (or column) commute and that the product of rows (or columns) yield $\hat{R}_{i}=\mathbb{I}$ and $\hat{C}_{j}=\mathbb{I}\,(j\neq3)$, $\hat{C}_{3}=-\mathbb{I}$, ($\forall\,i,j$) where $\hat{R}_{i}\equiv\prod_{j}\hat{A}_{ij}$, $\hat{C}_{j}\equiv\prod_{i}\hat{A}_{ij}$. Let us assume that $m$ exists. From property (b) of the map, to get $m(\hat{C}_{3})=-1$ (as required by property (a)), we must have an odd number of $-1$ assignments in the third column. In the remaining columns, the number of $-1$ assignments must be even (for each column). Thus, in the entire square, the number of $-1$ assignments must be odd. Let us use the same reasoning, but along the rows. Since each $m(\hat{R}_{i})=1$, we must have even number of $-1$ assignments along each row. Thus, in the entire square, the number of $-1$ assignments must be even. We have arrived at a contradiction and therefore we conclude that our assumption that a consistent $m$ exists, must be wrong.
\end{proof}

\begin{rem}
One could in principle assume $m$, to be s.t. (a) $m(\hat{\mathbb{I}})=1$,
(b) $m(\alpha\hat{B}_{i})=\alpha m(\hat{B}_{i})$, for $\alpha\in\mathbb{R}$,
(c) $m(\hat{B}_{i}^{2})=m(\hat{B}_{i})^{2}$, (d) $m(\hat{B}_{i}+\hat{B}_{j})=m(\hat{B}_{i})+m(\hat{B}_{j})$,
to deduce (d') $m(\hat{B}_{i}\hat{B}_{j})=m(\hat{B}_{i})m(\hat{B}_{j})$
and that $m(\hat{B}_{i})\in$ spectrum of $\hat{B}_{i}$. Effectively
then, condition (b) listed in the theorem is satisfied as a consequence.
Therefore, assuming (a)-(d) as listed here, rules out a larger class
of $m$. \cite{KochenSpecker}
\end{rem}
According to \thmref{KS}, non-contextual maps which are \emph{multiplicative}
must be incompatible with QM. It is apparent that non-contextual maps
which are non-\emph{multiplicative} might be consistent with QM. To
that end, in the following section we first show that this is indeed
the case by constructing an explicit model and supplement it by removing contextuality from Bohmian Mechanics.

>>>>

\documentclass[british,aps,prl,superscriptaddress,nofootinbib,times,reprint]{revtex4-1}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm,headheight=2cm,headsep=2cm,footskip=2cm}
\setcounter{secnumdepth}{3}
\usepackage{verbatim}
\usepackage{refstyle}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{esint}
\makeatletter
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
\AtBeginDocument{\providecommand\thmref[1]{\ref{thm:#1}}}
\AtBeginDocument{\providecommand\lemref[1]{\ref{lem:#1}}}
\AtBeginDocument{\providecommand\propref[1]{\ref{prop:#1}}}
\AtBeginDocument{\providecommand\eqref[1]{\ref{eq:#1}}}
\AtBeginDocument{\providecommand\figref[1]{\ref{fig:#1}}}
\AtBeginDocument{\providecommand\secref[1]{\ref{sec:#1}}}
\AtBeginDocument{\providecommand\ssecref[1]{\ref{ssec:#1}}}
\AtBeginDocument{\providecommand\defnref[1]{\ref{defn:#1}}}
\AtBeginDocument{\providecommand\tblref[1]{\ref{tbl:#1}}}

\newref{tbl}{name=Table~}{}
\newref{defn}{name=Definition~}{}
\newref{sec}{name=Section~}
\newref{ssec}{name=Subsection~}

@@

\usepackage{babel}
  \providecommand{\definitionname}{Definition}
  \providecommand{\examplename}{Example}
  \providecommand{\lemmaname}{Lemma}
  \providecommand{\propositionname}{Proposition}
  \providecommand{\remarkname}{Remark}
  \providecommand{\theoremname}{Theorem}
  \providecommand{\notationname}{Notation}
\setlength{\itemindent}{0pt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%This stuff is required for transposing matrices | fine I didn't want to do donkey work
\usepackage{xparse}
\usepackage{environ}

\ExplSyntaxOn
\NewEnviron{bmatrixT}
 {
  \marine_transpose:V \BODY
 }

\int_new:N \l_marine_transpose_row_int
\int_new:N \l_marine_transpose_col_int
\seq_new:N \l_marine_transpose_rows_seq
\seq_new:N \l_marine_transpose_arow_seq
\prop_new:N \l_marine_transpose_matrix_prop
\tl_new:N \l_marine_transpose_last_tl
\tl_new:N \l_marine_transpose_body_tl

\cs_new_protected:Nn \marine_transpose:n
 {
  \seq_set_split:Nnn \l_marine_transpose_rows_seq { \\ } { #1 }
  \int_zero:N \l_marine_transpose_row_int
  \prop_clear:N \l_marine_transpose_matrix_prop
  \seq_map_inline:Nn \l_marine_transpose_rows_seq
   {
    \int_incr:N \l_marine_transpose_row_int
    \int_zero:N \l_marine_transpose_col_int
    \seq_set_split:Nnn \l_marine_transpose_arow_seq { & } { ##1 }
    \seq_map_inline:Nn \l_marine_transpose_arow_seq
     {
      \int_incr:N \l_marine_transpose_col_int
      \prop_put:Nxn \l_marine_transpose_matrix_prop
       {
        \int_to_arabic:n { \l_marine_transpose_row_int }
        ,
        \int_to_arabic:n { \l_marine_transpose_col_int }
       }
       { ####1 }
     }
   }
   \tl_clear:N \l_marine_transpose_body_tl
   \int_step_inline:nnnn { 1 } { 1 } { \l_marine_transpose_col_int }
    {
     \int_step_inline:nnnn { 1 } { 1 } { \l_marine_transpose_row_int }
      {
       \tl_put_right:Nx \l_marine_transpose_body_tl
        {
         \prop_item:Nn \l_marine_transpose_matrix_prop { ####1,##1 }
         \int_compare:nF { ####1 = \l_marine_transpose_row_int } { & }
        }
      }
     \tl_put_right:Nn \l_marine_transpose_body_tl { \\ }
    }
   \begin{bmatrix}
   \l_marine_transpose_body_tl
   \end{bmatrix}
 }
\cs_generate_variant:Nn \marine_transpose:n { V }
\cs_generate_variant:Nn \prop_put:Nnn { Nx }
\ExplSyntaxOff

\begin{document}
\title{A non-contextual hidden variable model for quantum mechanics}
\author{Atul Singh Arora}
\email{ms11003@iisermohali.ac.in}
\selectlanguage{british}%
\affiliation{Indian Institute of Science Education and Research (IISER), Mohali}
\author{Arvind}
\email{arvind@iisermohali.ac.in}
\selectlanguage{british}%
\affiliation{Indian Institute of Science Education
and Research (IISER), Mohali}
\begin{abstract}
We propose a non-contextual hidden variable model,
consistent with all predictions of quantum
mechanics (QM).  A careful scrutiny of
consistency requirements between any hidden
variable model and  quantum mechanics and the
corresponding no-go theorems, leads us to the
conclusion that the notion of contextuality is not
a necessary feature of QM.  The alternative view
that emerges, hinges on identifying a new classical
property, which we call ``multiplicativity''.  It turns
out that, by relaxing this condition of
``multiplicativity'' on hidden variable models, they
can be made  consistent with QM.  Advantages of
this view are illustrated by considering the
implications of non-multiplicativity to locality,
entanglement, superposition, and  apparent
contextuality. 
\end{abstract}
\pacs{}
\maketitle

Quantum mechanics has been one of the most
successful theories in physics so far, however,
there has not yet been a final word on its
completeness and interpretation~\cite{??}.
Einstein's~\cite{EinsteinEPR} work on the
incompleteness of QM and the subsequent seminal
work of Bell~\cite{Bell1964}, assessing the
compatibility of a more complete model involving
hidden variables (HV) and locality with QM, has
provided deep insights into  how the quantum world
differs from its classical counterpart.  In recent
times, these insights have been of pragmatic
utility in the areas of quantum information
processing, where EPR pairs are fundamental motifs
of entanglement~\cite{Ekert,PironioRndmnssCrtfcn}.
The work of  Kochen Specker~\cite{KochenSpecker}
and Gleason et.
al.~\cite{Gleason,BellOnHiddenVariables,Peres,Mermin})
broadened the schism between  HV models and QM.
They showed that it was contextuality and not
non-locality which was at the heart of this schism
and the incompatibility between HV models and QM
can arise even for a single quantum system.
Contextuality has thus been identified as a
fundamental non-classical feature of the quantum
world and experiments have also been proposed and
conducted to this effect~\cite{SimonContExpProp,
HuangContExp,YangContExp,HasegawaContExp}.
Contextuality, on the one hand has led to
investigations on the foundational aspects of
QM~\cite{PawelCntxClsscl,?????}, and on the other
hand has been harnessed for computation and
cryptography~\cite{HowardCntxCmptn,CabelloCntxScrt,??}.
Quantitative measures of contextuality based on
memory~\cite{MatthiasCntxMmry} have also been used
to demonstrate the completeness of
QM~\cite{CabelloMmryQM}.  However, not all HV
models are incompatible with QM, and a very
important case in point is  Bohm's model based on
precise trajectories for quantum
systems~\cite{Bohm1,Bohm2}.

A closer look at the aforesaid no-go theorems
reveals that contextuality is in fact not a
necessary feature of QM (see \figref{block}).
While this has been shown earlier by taking
disturbances due to measurements into
consideration~\cite{NoContextuality,LaCourNoCntx},
we take a different approach and construct a
HV model consistent with QM.  The predictions of
the aforesaid model match with those of QM for systems
described by arbitrary finite-dimensional Hilbert
spaces.  \begin{figure}[h]
\includegraphics[width=0.9\columnwidth]{block1}
\caption{Non-contextuality is not inconsistent
with QM} \label{fig:block}\end{figure} The
algebraic constraints obeyed by quantum
observables may not necessarily be obeyed by the
predictions of an HV model at the level of
individual outcomes.  Those HV models that obey
such algebraic constraints possess a special
property which we call `multiplicativity'.  
Enlarging the scope of HV models to include
non-multiplicative models, permits us to construct an
explicit non-contextual model consistent with QM.
The re-examination of the very proof of
`contextuality'  is used to illustrate
non-multiplicativity, an intrinsic quantum feature, 
which we propose as a possible alternative to contextuality.

{\noindent \bf Notation:} (a) $\psi\in\mathcal{H}$ 
represents a pure quantum mechanical
state of the system in the Hilbert space
$\mathcal{H}$, (b) $\hat{\mathcal{H}}$ is defined
to mean the set of Hermitian 
observables for the system, (c)
$[\mathcal{H}]$ is defined to mean
$(\mathcal{H},\,\mathbb{R}^{\otimes})$, which
represents the state of the system including HVs,
(d) $[\psi]\in[\mathcal{H}]$ will represent the
state of the system, including HVs, (e) a
prediction map is
$m:\hat{\mathcal{H}},[\mathcal{H}]\to\mathbb{R}$,
(f) a sequence map is
$s:\hat{\mathcal{H}},[\mathcal{H}],\mathbb{R}\to[\mathcal{H}]$,
(g) $f$ is an arbitrary map from $\{
\hat{\mathcal{H}},\hat{\mathcal{H}},\dots\hat{\mathcal{H}}
\} \to \hat{\mathcal{H}}$ constructed using
multiplication and addition of the observables,
and multiplication with complex
numbers\footnote{Strictly, the map $f$ would
depend on the operators to ensure preservation of
Hermiticity; thus a given $f$ will be defined only
for a subset of $\{
\hat{\mathcal{H}},\hat{\mathcal{H}},\dots\hat{\mathcal{H}}
\}$.}, (h) $\tilde{f}$ is a map constructed by
replacing observables in $f$ with real numbers.

\begin{defn} A theory is non-contextual, if it
provides a map $m: \{
\hat{\mathcal{H}},[\mathcal{H}] \} \to\mathbb{R}$
to explain measurement outcomes. A theory which is
not non-contextual is contextual.\end{defn}

\begin{defnrem} The notion of non-contextuality may be
extended to maps. A map is non-contextual if it is
of the form $m: \{ \hat{\mathcal{H}},[\mathcal{H}]
\} \to\mathbb{R}$; prediction maps {\color{red}Comment: Is this really needed ``for individual
observables''} are non-contextual by construction.  
\end{defnrem}
\begin{defnrem}
Broader definitions in the literature have been
suggested which declare a larger set of theories
as non-contextual. For our purposes however, this
restricted definition will suffice. The term
contextual is used to suggest that the value an
operator takes might depend on which other
compatible observable it is being  measured with.
\end{defnrem}

Before we re-examine the result of Kochen and
Specker~\cite{KochenSpecker} that non-contextual
models cannot be compatible with quantum
mechanics and show that multiplicativity can provide an
alternative to contextuality, we define
multiplicativity more precisely.
\begin{defn}
A prediction map $m$ is \emph{multiplicative} iff
\begin{align*} &
m(f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{N}),[\psi])
= \\ & \tilde
f(m(\hat{B}_{1},[\psi]),m(\hat{B}_{2},[\psi]),\dots
m(\hat{B}_{N},[\psi])), \end{align*}
where $\hat{B}_{i}\in\hat{\mathcal{H}}$ are
arbitrary mutually commuting observables and
$[\psi]\in[\mathcal{H}]$. A
\emph{non-multiplicative} map is one that is not
multiplicative.  
\end{defn} 
Note that if $m$ is taken to represent the
measurement outcome (in QM), then for states of
the system which are simultaneous eigenkets of
$\hat{B}_{i}$s, $m$ must clearly be
multiplicative. It is, however, not obvious that
this property must always hold. For example,
consider two spin-half particles in the state 
$\left|1\right\rangle \otimes\left|1\right\rangle
$ written in the computational basis and take
operators 
$\hat{B}_{1}=\hat{\sigma}_{x}\otimes\hat{\sigma}_{x}$,
$\hat{B}_{2}=\hat{\sigma}_{y}\otimes\hat{\sigma}_{y}$
and
$\hat{C}=\hat{B}_{1}\hat{B}_{2}=-\hat{\sigma}_{z}\otimes\hat{\sigma}_{z}$
written in terms of Pauli operators.
We must have $m(\hat{C})=-1$ while
$m(\hat{B}_{1})=\pm1$ and $m(\hat{B}_{2})=\pm1$
independently, according to QM, with probability
half. Here multiplicativity does not seem to hold.
Antithetically, it is clear that if one first
measures $\hat{B}_{1}$ and subsequently measures
$\hat{B}_{2}$, then the product of the results
must be $-1$. This is consistent with measuring
$\hat{C}$. To capture this property we define
\emph{sequential multiplicativity} as follows.
\begin{defn} 
A prediction map $m$ is
\emph{sequentially multiplicative} for a given
sequence map $s$, iff \begin{align*}
&m(f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{N}),[\psi_{1}])=\\
&\tilde
f(m(\hat{B}_{1},[\psi_{k_{1}}]),m(\hat{B}_{2},[\psi_{k_{2}}]),\dots,m(\hat{B}_{N},[\psi_{k_{N}}])),
\end{align*} 
where
$\mathbf{k}=(k_{1},k_{2},\dots,k_{N})\in\{
N! {\rm ~permutations~of~}  k'{\rm s}
\}$, $\hat{B}_{i}\in\hat{\mathcal{H}}$ are
arbitrary mutually commuting observables,
$[\psi_{i}]\in[\mathcal{H}]$ and
$[\psi_{k+1}] :=
s(\hat{B}_{k},[\psi_{k}],m(\hat{B}_{k},[\psi_{k}]))$,
$\forall\,[\psi_{i}]$.  
\label{defn:seqnMltpl}\end{defn} 
With these definitions we are now ready to discuss
the `proof of contextuality'. We first state the
contextuality theorem in our notation:
\begin{thm} Let a map
$m:\hat{\mathcal{H}}\to\mathbb{R}$, be s.t. (a)
$m(\hat{\mathbb{I}})=1$, (b)
$m(f(\hat{B}_{1},\hat{B}_{2},\dots))=\tilde
f(m(\hat{B}_{1}),m(\hat{B}_{2}),\dots)$, for any
arbitrary function $f$, where $\hat{B}_{i}$ are
mutually commuting Hermitian operators. If $m$ is
assumed to describe the outcomes of measurements,
then no $m$ exists which is consistent with all
predictions of QM. 
\label{thm:KS}
\end{thm}

\begin{proof} Peres Mermin
($\left|\mathcal{H}\right|\ge4$)
\cite{Peres,Mermin}:
For a system composed of two spin-half particles 
consider the following set of operators \[
\hat{A}_{ij}\doteq\left[\begin{array}{ccc}
\hat{\mathbb{I}}\otimes\hat{\sigma}_{x} &
\hat{\sigma}_{x}\otimes\hat{\mathbb{I}} &
\hat{\sigma}_{x}\otimes\hat{\sigma}_{x}\\
\hat{\sigma}_{y}\otimes\hat{\mathbb{I}} &
\hat{\mathbb{I}}\otimes\hat{\sigma}_{y} &
\hat{\sigma}_{y}\otimes\hat{\sigma}_{y}\\
\hat{\sigma}_{y}\otimes\hat{\sigma}_{x} &
\hat{\sigma}_{x}\otimes\hat{\sigma}_{y} &
\hat{\sigma}_{z}\otimes\hat{\sigma}_{z}
\end{array}\right] \] which have the property that
all operators along a row (or column) commute. Further,
the product of rows (or columns) yields
$\hat{R}_{i}=\mathbb{I}$ and
$\hat{C}_{j}=\mathbb{I}\,(j\neq3)$,
$\hat{C}_{3}=-\mathbb{I}$, ($\forall\,i,j$) where
$\hat{R}_{i}:=\prod_{j}\hat{A}_{ij}$,
$\hat{C}_{j}:=\prod_{i}\hat{A}_{ij}$. Let us
assume that $m$ exists. From property (b) of the
map, to get $m(\hat{C}_{3})=-1$ (as required by
property (a)), we must have an odd number of $-1$
assignments in the third column. In the remaining
columns, the number of $-1$ assignments must be
even (for each column). Thus, in the entire
square, the number of $-1$ assignments must be
odd. Let us use the same reasoning, but along the
rows. Since each $m(\hat{R}_{i})=1$, we must have
even number of $-1$ assignments along each row.
Thus, in the entire square, the number of $-1$
assignments must be even. We have arrived at a
contradiction and therefore we conclude that our
assumption that a consistent $m$ exists, must be
wrong.  \end{proof}

\begin{thmrem} One could in principle assume $m$, to
be s.t. (a) $m(\hat{\mathbb{I}})=1$, (b)
$m(\alpha\hat{B}_{i})=\alpha m(\hat{B}_{i})$, for
$\alpha\in\mathbb{R}$, (c)
$m(\hat{B}_{i}^{2})=m(\hat{B}_{i})^{2}$, (d)
$m(\hat{B}_{i}+\hat{B}_{j})=m(\hat{B}_{i})+m(\hat{B}_{j})$,
to deduce (d')
$m(\hat{B}_{i}\hat{B}_{j})=m(\hat{B}_{i})m(\hat{B}_{j})$
and that $m(\hat{B}_{i})\in$ spectrum of
$\hat{B}_{i}$. Effectively then, condition (b)
listed in the theorem is satisfied as a
consequence.  Therefore, assuming (a)-(d) as
listed here, rules out a larger class of $m$.
\cite{KochenSpecker} 
\end{thmrem} 
Here $m$ maybe viewed as a specific
class of prediction maps, that implicitly depends
on the state $[\psi]$. It is clear that according to
\thmref{KS}, non-contextual maps which are
\emph{multiplicative} must be incompatible with
QM. However, non-contextual maps which are
non-\emph{multiplicative} might be consistent with
QM. Indeed it is and we will show this by constructing an
explicit model.
